taxadb: A High-Performance Local Taxonomic Database Interface

### 2019-04-03 paper/manuscript.Rmd
WIP
pdf

journal: "Methods in Ecology & Evolution"
date: "`r Sys.Date()`"
bibliography: refs.bib
layout: 3p
output: rticles::elsevier_article
---

```{r message=FALSE, include = FALSE}

@@

library(printr)
knit_print.data.frame <- function(x, ...) {
#  knitr::knit_print(DT::datatable(x), ...)
  knitr::knit_print(kableExtra::kable(x, "latex"), ...)
}

knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE)
```

@@

```{r }
status_data
```

```{r}
trait_data
```

@@

```{r}
full_join(trait_data, status_data, by = c("name_A" = "name_B"))
```

If we first resolve names used in each data set into shared identifiers, (for instance, using the Catalogue of Life), we discover that there is far more overlap in the species coverage than we might have realized at first.

@@

traits <- trait_data %>% mutate(id = get_ids(name_A, "col"))
status <- status_data %>% mutate(id = get_ids(name_B, "col"))

full_join(traits, status, by = "id") %>% 
  select(id, name_A, mass, category, name_B) # just reordering

```

@@

---

Box: `get_ids()` returns the `acceptedNameUsageID`, the identifier associated with the *accepted* name.  Some naming providers, such as ITIS and NCBI, provide taxonomic identifiers to both synonyms and accepted names.  Other providers, such as COL and GBIF, only provide identifiers for accepted names.  
Common practice in Darwin Core archives is to provide an `acceptedNameUsageID` only for names which are synonyms, and otherwise to provide a `taxonID`.  For accepted names, the `acceptedNameUsageID` is then given as missing (`NA`), while for synonyms, the `taxonID` may be missing (`NA`).  In contrast, `taxadb` lists the `acceptedNameUsageID` for accepted names (where it matches the `taxonID`), as well as known synonyms.  This is semantically identical, but also more convenient for database interfaces, since it allows a name to mapped to it's accepted identifier (or an identifier to map to it's accepted name usage) without the additional logic.  For consistency, we will use the term "identifier" to mean the `acceptedNameUsageID` rather than the more ambiguous `taxonID` (which is undefined for synonyms listed by many providers), unless explicitly stated otherwise.

---

@@

Walking synonyms between providers should be done with care.  Synonyms

An accepted name can have many synonyms:

@@

>>>>

journal: "Methods in Ecology & Evolution"
date: "`r Sys.Date()`"
bibliography: refs.bib
layout: 3p
output: 
  rticles::springer_article:
    includes:
      in_header: preamble.tex
---

```{r message=FALSE, include = FALSE}
library(kableExtra)
library(magrittr)
library(tidyverse)
library(taxadb)

@@

knitr::opts_chunk$set(cache=TRUE, message=FALSE, warning=FALSE)
```

@@

```{r echo=FALSE, cache = FALSE}
status_data %>% kableExtra::kable("latex", booktabs=T) %>% 
                      kable_styling(latex_options = "striped", full_width = F)
```

```{r echo = FALSE, cache = FALSE }
trait_data %>% kableExtra::kable("latex", booktabs=T) %>% 
                      kable_styling(latex_options = "striped", full_width = F)
```

@@

```{r}
joined <- full_join(trait_data, status_data, by = c("elton_name" = "redlist_name"))
```

```{r echo = FALSE, cache = FALSE}
joined %>% kableExtra::kable("latex", booktabs=T) %>% 
                      kable_styling(latex_options = "striped", full_width = F)
```

If we first resolve names used in each data set into shared identifiers, (for instance, using the Catalogue of Life), we discover that there is far more overlap in the species coverage than we might have realized at first.  First, we just add an ID column to each table by looking up the Catalog of Life identifier for the name provided:

@@

traits <- trait_data %>% mutate(id = get_ids(elton_name, "col"))
status <- status_data %>% mutate(id = get_ids(redlist_name, "col"))
```

We can now join on the `id` column instead of names directly:

```{r}
joined <- full_join(traits, status, by = "id") 
```

```{r cache = FALSE, echo = FALSE}
joined %>%  
  select(elton_name, redlist_name, mass, category, id) %>%
  kableExtra::kable("latex", booktabs=T) %>% 
  kable_styling(latex_options = "striped", full_width = F)
```

@@

\dummy{\Begin{tcolorbox}[title= Box 1: Taxonomic Identifiers and Synonyms, lower separated=false]}

`get_ids()` returns the `acceptedNameUsageID`, the identifier associated with the *accepted* name.  Some naming providers, such as ITIS and NCBI, provide taxonomic identifiers to both synonyms and accepted names.  Other providers, such as COL and GBIF, only provide identifiers for accepted names.  Common practice in Darwin Core archives is to provide an `acceptedNameUsageID` only for names which are synonyms, and otherwise to provide a `taxonID`.  For accepted names, the `acceptedNameUsageID` is then given as missing (`NA`), while for synonyms, the `taxonID` may be missing (`NA`).  In contrast, `taxadb` lists the `acceptedNameUsageID` for accepted names (where it matches the `taxonID`), as well as known synonyms.  This is semantically identical, but also more convenient for database interfaces, since it allows a name to mapped to it's accepted identifier (or an identifier to map to it's accepted name usage) without the additional logic.  For consistency, we will use the term "identifier" to mean the `acceptedNameUsageID` rather than the more ambiguous `taxonID` (which is undefined for synonyms listed by many providers), unless explicitly stated otherwise.

\End{tcolorbox}

@@

## Synonyms

Walking synonyms between providers should be done with care.  An accepted name can have many synonyms:

@@

\pagebreak

# References
